PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-02-13 papers/NAR_web_server_2009/paper.tex
Several changes to the paper to actually make it more LaTex friendly. One of the important modifications is to change Figure~1 to Figure~\ref{fig:workflow}. Figures need to be referred. Otherwise links and layout will be broken.
I have done these references only to a few spots and we need to scan the 
paper for MS-Word like use of Figure~2(c) etc. and replace them with LaTex
references.

I have also replaced all uses of {\sc peace} with custom macro \peace which
will permit us to conveniently refer and change the formatting by touching
just 1 line.

I cropped several pdf files with pdfcrop to

\documentclass[a4,center,fleqn]{NAR}
\usepackage[pdftex,
colorlinks=true, %makes the links show up in color, rather than in a box
citecolor=black, %color of the in-text citation numbers
urlcolor=blue %color of the url links
]{hyperref}

@@

%\articlesubtype{This is the article type (optional)}

@@

\begin{abstract}
  We present {\sc Peace}, a stand alone tool for the high-throughput
  {\it ab initio} clustering of transcript fragment sequences produced
  by Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, {\sc Peace} can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tools.  Through the GUI the user can collect
  cluster statistics and examine specific clusters for more
  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
  enables any assembly tool to work with individual small clusters of
  closely related sequences, resulting in significant increases in
  sensitive and robustness to sequencing error, as well as a reduction in
  overall runtime.
\end{abstract}

@@

But clustering is a computationally challenging problem.  Even with
the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster based on pair-wise sequence
alignments make such an approach infeasible in practice.  The much
larger data-set size produced by NGS technologies exacerbates this
problem.  To deal with this, {\sc Peace} combines the $d^2$
alignment-free sequence distance function \cite{Hide94} and the
concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
accurately find clusters of ESTs expressed from the same gene without
reference to a sequenced genome.  Compared against competing tools
for the same purpose \cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Ptitsyn05,Hazelhurst08a,Picardi09}, {\it Peace} proves to be both more sensitive and
more robust to sequencing error without sacrificing runtime.

In short, {\sc Peace} is computational tool for the {\it ab initio}
clustering of transcript fragments by gene association, applicable to both NGS and
traditional Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
{\sc Peace} GUI allows the user to both easily install (locally or
remotely) and run the clustering engine, as well as enabling
transparent parallel processing and providing various tools for result
analysis.

@@

\section{{\sc Peace}: Installation and Use}

The {\sc Peace} GUI can be launched by downloading and executing the
Jar file available on the {\sc Peace} website
(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
machine running the standard Java Virtual Machine (JVM).  Once running,
the user can employ the GUI to install the computational tool and perform
a clustering of a data file in the FASTA format, view an initial
analysis of the clusters, and produce files containing subsets of the
clusters as input to assembly tools such as {\sc Cap3} \cite{Huang99}.
A typical (first) use of {\sc Peace} generally
requires the following steps.

\noindent {\bf Tool Installation:} (First use only.)  To install the
{\sc Peace} clustering onto a local or remote machine, the
user selects from within the GUI the appropriate menu tab
(Figure~1(a)), which then starts an install wizard that will prompt
for the appropriate information.  Figure~1(b) illustrates the request
for server information; the user has chosen to install the {\sc Peace}
computational tool on a remote machine and is providing the necessary
connection information.  Server information is persistent between GUI
sessions, giving the user access to {\sc Peace} on the target machine
as needed.

\noindent {\bf Job Processing:} After importing the target sequence file into
the GUI, the user starts a new job by following the wizard menus.
Figure~1(c) illustrates the process of specifying the number of
processors available (if running on a machine supporting the OpenMPI
protocol -- which will be determined during job installation).  Once
executed, the GUI will manage the job thread, alert the user when the
job is completed (or when the user next runs the GUI after completion),
and copy the final results back to the local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original target file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences (illustrated
  in Figure~1(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user can set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  ware matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential expression profiles
  of different libraries for a given cluster.  The method of setting up these classifiers, and the
  resulting histogram, is illustrated in Figure~1(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos demonstrating
{\sc Peace} use and capabilities.

@@

The clustering performed by {\sc Peace} is based on the use of minimum
spanning trees (MSTs), known to be an effective approach for narrow
band single linkage clustering \cite{Jain99,Wan08}. Edge weights are
assigned using the $d^2$ distance measure \cite{Hide94} to represent
edge weights.  Prim's algorithm \cite{Prim57} then allows us to
efficiently calculate an MST from which we can infer a high-quality
clustering solution.

@@

{\sc Peace} approaches the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements) and can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~A and B of the Supplementary Materials for more details).

@@

{\sc Peace} has been test on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
an implicit, intermediate step).  For our simulation tests we used the
{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
of varying length fragments under different models of error
(Supplementary Materials, Section~C.1), generating the fragments from
the list of 100 zebra fish genes used in the {\sc wcd} testing
\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
as possible, those used in the {\sc wcd} study (see supplementary
materials).  Our primary methods of quality assessment were {\it
  sensitivity} (the fraction of fragment pairs from the same gene that
were correctly clustered together) and {\it Type 1 error} (the
fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
the supplementary materials.  In Figure~2(left) we see a significant
improvement of {\sc Peace} (blue) over {\sc wcd} (green) and {\sc
  Cap3} (black) in its sensitivity to sequencing errors, while in
Figure~2(right) we see a comparable improvement in Type 1 error.  In
Figure~3 we look at the number of singleton clusters (fragments not
joined to any cluster), which will not occur in our simulated sets; we
again see significant improvement in {\sc Peace}.  The sequential
runtime of {\sc Peace} is slightly improved over that of {\sc wcd}
(see Supplementary Materials, Figure~S4).

\begin{figure*}[b]
\centerline{
\includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf} 
}
\label{screen}
\caption{Screenshots of the {\sc Peace} GUI during execution, including (a)
GUI Welcome and server installation menu; (b) setup wizard for
installing the computational tool on a remote server; (c) execution
wizard for starting a selected job to be executed in parallel mode;
(d) basic cluster output; and (e) histogram view of cluster results
and classifier editor for setting up differential expression profiles.}
\end{figure*}

\begin{figure*}[h]
\centerline{
\begin{minipage}[b]{2in}
\includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
\end{minipage}{2in}
}
\label{SeT1}
\caption{Comparisons of Sensitivity and Type 1 error, based on the
  average over 30 simulated fragment sets derived from 100 zebra fish genes
  (see Supplementary Materials, Section~\href{sim_results}, for more
  details).   Blue/Solid = {\sc Peace}, Green/Dash =
  {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
  intervals on estimates.  Intervals are not presented for Type 1
  error due to the extremely small variance.}
\end{figure*}

\begin{figure*}[b]
\begin{minipage}{2in}
\centerline{
\includegraphics[scale=0.35]{pics.d/singletons.pdf}
}
\end{minipage}
\label{singletons}
\caption{Average number of fragments flagged as singletons by each tool
  when run on the simulated sequences; correct answer in all cases is
  zero.  Blue/Solid = {\sc Peace}, Green/Dash = {\sc wcd},
  Black/Dot-Dash = {\sc Cap3}.}
\end{figure*}

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in {\sc Peace} over {\sc wcd} in both
sensitivity and Type 1 error (both significantly outperforming {\sc
  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
\cite{Hazelhurst08a} {\sc Peace} again shows slight improvements in
sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
  Peace} (see Supplementary Materials, Section~D, for more details).

@@

Here we have presented {\sc Peace}, a stand alone tool for the
high-throughput clustering of transcript fragments, managed through a
user friendly GUI that enables both local and remote installation and
execution in sequential or parallel mode.  Based on a novel algorithm
for the clustering of the fragments by gene association, {\sc Peace}
shows significant improvement in sensitivity, without sacrificing runtime,
over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
order of magnitude in improvement over the clustering preformed in the
course of assembly by the {\sc Cap3} tool \cite{Huang99}.

As a clustering tool based on sequence distance, {\sc Peace} faces
certain inherent limitations. For example, {\sc Peace} cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, natural biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a whole)
results in a significant reduction in overall assembly time.

@@

\vspace{3mm}
%Bib TeX

>>>>

\documentclass[a4,center,fleqn]{NAR}
\usepackage[pdftex,
colorlinks=true, %makes the links show up in color, rather than in a box
citecolor=black, %color of the in-text citation numbers
urlcolor=blue %color of the url links
]{hyperref}
\usepackage{wrapfig}

@@

% \articlesubtype{This is the article type (optional)}

% Some shortcuts to make life easier when writing latex
\newcommand{\peace} {{\small PEACE}}

@@

\begin{abstract}
  We present \peace, a stand alone tool for the high-throughput {\it
    ab initio} clustering of transcript fragment sequences produced by
  Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, \peace\/ can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tools.  Through the GUI the user can collect cluster
  statistics and examine specific clusters for more comprehensive
  study.  \peace\/, used as a ``pre-assembly'' tool, enables any
  assembly tool to work with individual small clusters of closely
  related sequences, resulting in significant increases in sensitive
  and robustness to sequencing error, as well as a reduction in
  overall runtime.
\end{abstract}

@@

But clustering is a computationally challenging problem.  Even with
the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster based on pair-wise sequence
alignments make such an approach infeasible in practice.  The much
larger data-set size produced by NGS technologies exacerbates this
problem.  To deal with this, \peace\/ combines the $d^2$
alignment-free sequence distance function \cite{Hide94} and the
concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
accurately find clusters of ESTs expressed from the same gene without
reference to a sequenced genome.  Compared against competing tools for
the same purpose
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
sensitive and more robust to sequencing error without sacrificing
runtime.

In short, \peace\/ is computational tool for the {\it ab initio}
clustering of transcript fragments by gene association, applicable to
both NGS and traditional Sanger Sequencing technologies.  Available
through the \href{http://www.peace-tools.org}{www.peace-tools.org}
website, the \peace\/ GUI allows the user to both easily install
(locally or remotely) and run the clustering engine, as well as
enabling transparent parallel processing and providing various tools
for result analysis.

@@

\section{\peace\/: Installation and Use}

The \peace\/ GUI can be launched by downloading and executing the
Jar file available on the \peace\/ website
(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
machine running the standard Java Virtual Machine (JVM).  Once
running, the user can employ the GUI to install the computational tool
and perform a clustering of a data file in the FASTA format, view an
initial analysis of the clusters, and produce files containing subsets
of the clusters as input to assembly tools such as {\sc Cap3}
\cite{Huang99}.  A typical (first) use of \peace\/ must be
performed in the following manner (see Figure~\ref{fig:workflow}):

\begin{figure}
  \centerline{\includegraphics[width=3in]{screen.d/workflow-crop.pdf}}
  \caption{Overview of the procedure for clustering and analysis using
    PEACE.}\label{fig:workflow}
\end{figure}

\noindent {\bf Tool Installation:} (First use) To install the {\sc
  Peace} clustering onto a local or remote machine, the user selects
from within the GUI the appropriate menu tab (Figure~\ref{screen}(a)),
which then starts an install wizard that will prompt for the
appropriate information.  Figure~\ref{screen}(b) illustrates the
request for server information; the user has chosen to install the
\peace\/ computational tool on a remote machine and is providing
the necessary connection information.  Server information is
persistent between GUI sessions, giving the user access to \peace\/
on the target machine as needed.

\begin{figure*}
  \centerline{ 
    \includegraphics[trim=0cm 3cm 0cm 1cm, clip, scale=0.9]{screen.d/screen.pdf}
  }
  \caption{Screenshots of the \peace\/ GUI during execution, including (a)
    GUI Welcome and server installation menu; (b) setup wizard for
    installing the computational tool on a remote server; (c) execution
    wizard for starting a selected job to be executed in parallel mode;
    (d) basic cluster output; and (e) histogram view of cluster results
    and classifier editor for setting up differential expression profiles.}\label{screen}
\end{figure*}

\noindent {\bf Job Processing:} After importing the target sequence
file into the GUI, the user starts a new job by following the wizard
menus.  Figure~\ref{screen}(c) illustrates the process of specifying
the number of processors available (if running on a machine supporting
the OpenMPI protocol -- which will be determined during job
installation).  Once executed, the GUI will manage the job thread,
alert the user when the job is completed (or when the user next runs
the GUI after completion), and copy the final results back to the
local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the
  original target file containing the sequences corresponding to the
  selected clusters ready for processing by an assembly tool
  (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences
  (illustrated in Figure~\ref{screen}(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user can set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  ware matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential
  expression profiles of different libraries for a given cluster.  The
  method of setting up these classifiers, and the resulting histogram,
  is illustrated in Figure~\ref{screen}(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos
demonstrating \peace\/ use and capabilities.

@@

The clustering performed by \peace\/ is based on the use of minimum
spanning trees (MSTs), known to be an effective approach for narrow
band single linkage clustering \cite{Jain99,Wan08}. Edge weights are
assigned using the $d^2$ distance measure \cite{Hide94} to represent
edge weights.  Prim's algorithm \cite{Prim57} then allows us to
efficiently calculate an MST from which we can infer a high-quality
clustering solution.

@@

\peace\/ approaches the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements) and can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~A and B of the Supplementary Materials for more details).

@@

\peace\/ has been test on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the {\sc wcd} clustering tool \cite{Hazelhurst08a} and the
{\sc Cap3} assembly tool \cite{Huang99} (which calculates a cluster as
an implicit, intermediate step).  For our simulation tests we used the
{\bf ESTSim} tool \cite{Hazelhurst03} to generate simulated transcript
of varying length fragments under different models of error
(Supplementary Materials, Section~C.1), generating the fragments from
the list of 100 zebra fish genes used in the {\sc wcd} testing
\cite{Hazelhurst08a}.  Tool parameters were taken to match, as closely
as possible, those used in the {\sc wcd} study (see supplementary
materials).  Our primary methods of quality assessment were {\it
  sensitivity} (the fraction of fragment pairs from the same gene that
were correctly clustered together) and {\it Type 1 error} (the
fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}; other measurements are also discussed in
the supplementary materials.  In Figure~2(left) we see a significant
improvement of \peace\/ (blue) over {\sc wcd} (green) and {\sc
  Cap3} (black) in its sensitivity to sequencing errors, while in
Figure~2(right) we see a comparable improvement in Type 1 error.  In
Figure~3 we look at the number of singleton clusters (fragments not
joined to any cluster), which will not occur in our simulated sets; we
again see significant improvement in \peace.  The sequential
runtime of \peace\/ is slightly improved over that of {\sc wcd}
(see Supplementary Materials, Figure~S4).

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeT1.pdf}}
  \label{SeT1}
  \caption{Comparisons of Sensitivity and Type 1 error, based on the
    average over 30 simulated fragment sets derived from 100 zebra fish genes
    (see Supplementary Materials, Section~\href{sim_results}, for more
    details).   Blue/Solid = \peace, Green/Dash =
    {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
    intervals on estimates.  Intervals are not presented for Type 1
    error due to the extremely small variance.}
\end{figure}

\begin{figure}
  \centerline{\includegraphics[width=3in]{pics.d/singletons.pdf}}
  \label{singletons}
  \caption{Average number of fragments flagged as singletons by each tool
    when run on the simulated sequences; correct answer in all cases is
    zero.  Blue/Solid = \peace, Green/Dash = {\sc wcd},
    Black/Dot-Dash = {\sc Cap3}.}
\end{figure}

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in \peace\/ over {\sc wcd} in both
sensitivity and Type 1 error (both significantly outperforming {\sc
  Cap3}).  Using the mouse data set used in Hazelhurst {\it et al.}
\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
sensitivity and Type 1 error rates, with an 18\% speedup for {\sc
  Peace} (see Supplementary Materials, Section~D, for more details).

@@

Here we have presented \peace, a stand alone tool for the
high-throughput clustering of transcript fragments, managed through a
user friendly GUI that enables both local and remote installation and
execution in sequential or parallel mode.  Based on a novel algorithm
for the clustering of the fragments by gene association, \peace\/
shows significant improvement in sensitivity, without sacrificing runtime,
over the competing {\sc WCD} tool \cite{Hazelhurst08a}, and shows an
order of magnitude in improvement over the clustering preformed in the
course of assembly by the {\sc Cap3} tool \cite{Huang99}.

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, natural biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a whole)
results in a significant reduction in overall assembly time.

@@

\vspace{3mm}
% Bib TeX
